\section{Experiments}

\subsection{Experimental Setup}

\textbf{Dataset}

The datasets utilized for training and evaluation are constructed from a total of 400 dipeptide systems. To assess generalization to novel molecular graphs, we strictly withhold a test set comprising 20\% of the dipeptides; these unseen systems are entirely excluded from the training phase. For the remaining 80\% of the dipeptides, we process raw Molecular Dynamics (MD) trajectories, each containing \num{2e6} frames. We partition these in-distribution trajectories chronologically: the first half (\num{e6} frames) forms the training set to optimize the main model, while the second half serves as the validation set to monitor in-distribution generalization. Following established protocols, we operate under the assumption that a trajectory length of \num{e6} frames provides a sufficiently robust approximation of the Boltzmann distribution for dipeptide systems \citep{machine_learning_implicit_solvation} (important: frames alone not sufficient. only half the traj length of issnet).

\textbf{Metrics}

Prior work \citep{machine_learning_implicit_solvation} demonstrates that relying solely on static force prediction errors is insufficient for model selection, as the lowest mean squared error (MSE) on atomic forces does not strictly correlate with the best approximation of the target Boltzmann distribution. To rigorously evaluate the thermodynamic fidelity of the trained models, we conduct MD simulations using the framework detailed in Section \ref{sec:todo} \citep{ase_placeholder}, employing our models as surrogate calculators.

To ensure consistent comparisons, all baseline models and ablations are evaluated on the same randomly selected molecule from both the train and test sets. We execute 256 parallel MD replicas utilizing our simulation framework. Following \citet{machine_learning_implicit_solvation}, we initialize these parallel simulations by randomly sampling starting conformations from the all-atom ground-truth data of the respective system. This informed initialization facilitates a brief equilibration phase of 0.1 ns per replica. Subsequent production runs of 1 ns per replica aggregate to 256 ns of total effective simulation time, yielding \num{2.56e6} extracted frames.

To maintain parity across evaluations, both the model-generated and ground-truth trajectories are uniformly subsampled to \num{e6} frames. We then project these structural ensembles onto relevant low-dimensional collective variables—specifically, Ramachandran dihedral angles and \gls{tica} coordinates. Finally, we compute the \gls{kl} divergence and the Wasserstein distance between the predicted and ground-truth marginal distributions on these projections. These metrics provide a robust, quantitative measure of how closely the simulated ensembles approximate the true target Boltzmann distribution.

\subsection{Model Complexity Ablation}

We evaluate the impact of model capacity through a targeted ablation study across three representative scales of network complexity, detailed in Table \ref{tab:model_params}. This focused approach is motivated by prior work \citep{machine_learning_implicit_solvation}, which demonstrated that the force prediction accuracy of similar architectures is generally robust to structural hyperparameters, with performance being overwhelmingly dictated by the chosen atomic embeddings. Our empirical results corroborate this observation; scaling the architecture from a compact baseline ($\approx$ 97k parameters) to a highly parameterized network ($\approx$ 5M parameters) yielded no significant performance differences in either static force prediction (MSE) or dynamic thermodynamic fidelity (KLD). 

\begin{table}[h]
    \centering
    \caption{Representative model configurations used for our complexity ablation. The chosen architectures scale the hidden dimension ($h_{\text{dim}}$) and number of interaction blocks to evaluate capacity constraints, following \citep{machine_learning_implicit_solvation}. We used the \textit{TBG + full} embedding scheme, whose contribution to the total parameter size is negligible ($<$0.4\%).}
    \label{tab:model_params}
    \begin{tabular}{rrr}
        \toprule
        Hidden Dimension ($h_{\text{dim}}$) & Interaction Blocks & Total Parameters \\
        \midrule
        32 & 3 & 97,185 \\
        128 & 7 & 530,689 \\
        512 & 11 & 5,011,073 \\
        \bottomrule
    \end{tabular}
\end{table}

\subsection{Feature Embedding Ablation}


\subsection{Train Trajectory Length Ablation}

To ablate the effect of trajectory length on model performance, we slice the first $n$ samples of the trajectory. Because the trajectories were logged sequentially during simulation, this approach accurately simulates shorter physical trajectory lengths. To mitigate the confounding variable of dataset size—where longer trajectories inherently contain more training samples—we subsample each slice to match the frame count of the shortest trajectory length under investigation. This isolates the ablation to assess the quality of the thermodynamic sampling, as subsampling a longer, more decorrelated trajectory yields better statistical convergence with respect to the underlying Boltzmann distribution. This is further detailed in the discussion (todo ref this later).