\section{Related Work}
\label{sec:related_work}


To overcome the limitations of classical implicit models, recent developments have leveraged \gls{ml} to refine implicit solvation. \citet{machine_learning_implicit_solvation} were the first to apply \gls{ml} to this domain, employing a \gls{gnn} trained with a force-matching objective to learn solvent effects from explicit solvent simulation data. While their method achieved lower \gls{kld} on molecules like chignolin and alanine dipeptide compared to the baseline \gls{gb} model, its transferability across different systems remained untested.

Subsequent approaches have focused on improving generalization and accuracy. One strategy proposed by \cite{katzberger2024general, katzberger2025rapid} trains a \gls{nn} to predict the difference between implicit and explicit solvent results by modifying parameters within a \gls{gb} baseline while preserving computational efficiency. While successful in reproducing free-energy profiles of small organic molecules, this model has not been validated on larger molecules exceeding 700 Da. Another approach, potential contrasting \cite{Airas2023}, refines a pretrained SchNet potential without labeled force data by discriminating between realistic and noise-like conformations. Although it outperformed the \gls{gb} baseline on training proteins, its transferability to mutations with lower sequence identity and intrinsically disordered peptides proved limited. Furthermore, \citet{röcken2025predictingsolvationfreeenergies} introduced ReSolv, which combines vacuum pre-training on \gls{qm} data with solvent refinement for small, drug-like molecules.

\citet{charron2025navigating} trained a SchNet-based \gls{cg} model (CGSchNet) using variational force matching on explicit-solvent all-atom \gls{md} trajectories of 50 structurally diverse CATH protein domains. Similar to our approach, all solvent degrees of freedom are integrated out; however, the solute is additionally coarse-grained to five beads per residue. The model was evaluated on unseen fast-folding proteins with less than 40\% sequence similarity to the training set by comparing free-energy surfaces and native-contact fractions against all-atom references.


\citet{park2026scaling} examined the training objective itself, showing that mean force matching requires 50$\times$ fewer training samples and 87\% less total atomistic simulation time compared to standard force-matching objectives while achieving better accuracy on the potential of mean force for unseen proteins. 

\citet{klein2023timewarp} proposed an enhanced sampling method that uses a normalizing flow as a proposal distribution within a Markov chain Monte Carlo scheme targeting the Boltzmann distribution. Trained offline on \gls{md} trajectories, the flow learns to make large time-steps of $10^5$--$10^6$~fs and transfers to unseen small peptides of 2--4 amino acids at all-atom resolution, providing wall-clock acceleration of equilibrium sampling relative to standard \gls{md}.
